# 2. Regression and model validation

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using  


### The structure and dimensions of the data

```{r, echo=FALSE}
learning2014 <- read.csv(file = "/Users/mikko/Documents/GitHub/IODS-project/data/learning2014.csv", header = T, sep=",")
str(learning2014)
```

The data is a subset of all the variables from the data: "International survey of Approaches to Learning". The data consists of 166 rows (observations) and 7 columns (variables).
The variables are "gender","age","attitude", "deep", "stra", "surf" and "points". The variables "gender" (in years) and "age" (M=male, F=female) are self-explanatory. The variable "attitude" describes global attitude toward statistics (it is sum of ten variables). The variable "deep" describes deep approach (it is a mean of three variables). The variable "stra" describes strategic approach (it is a mean of two variables). The variable "surf" describes surface approach (it is a mean of three variables). The variable "points" is exam points.

### A graphical overview and summaries of variables in the data


```{r, echo=FALSE}
summary(learning2014)

library(ggplot2)
library(GGally)

# create a more advanced plot matrix with ggpairs()
p <- ggpairs(learning2014, mapping = aes(col=gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))

# draw the plot
p
```


There are almost twice as many answers from females (n = 110) than males (n = 56). Their age varies between 17 and 55 years, half of the participants being at least 22 years old.

### Regression model

I chose the variables "attitude", "stra" and "surf" as explanatory variables for the target variable "points". Here is the summary for the fitted model:

```{r, echo=FALSE}
# fit a linear model
my_model <- lm(points ~ attitude + stra + surf, data = learning2014)

# print out a summary of the model
summary(my_model)
```

According to the model the variables "stra" and "surf" do not have statistically significant relationship with the target (p-values 0.117 and 0.466). Since the variable "surf" has bigger p-value, let us remove that from the model and fit a new model without it. Here is the new fitted model summary:

```{r, echo=FALSE}
# fit a linear model
my_model <- lm(points ~ attitude + stra, data = learning2014)

# print out a summary of the model
summary(my_model)
```

Now the model summary shows that the variable "stra" does not have statistically significant relationship with the target (p-value is 0.089), so I create a new model without it. Here is the final model summary:

```{r, echo=FALSE}
# fit a linear model
my_model <- lm(points ~ attitude, data = learning2014)

# print out a summary of the model
summary(my_model)
```


