---
title: "4. Clustering and classification"
output: html_document
params:
  date: !r Sys.Date()
---

## 4. Clustering and classification

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```


This week we are using a data (included in the MASS package of R) called "Boston", which consists of variables to study *Housing Values in Suburbs of Boston*. The variables in the data are "per capita crime rate by town", "proportion of residential land zoned for lots over 25,000 sq.ft", "proportion of non-retail business acres per town", "Charles River dummy variable (= 1 if tract bounds river; 0 otherwise)", "nitrogen oxides concentration (parts per 10 million)", "average number of rooms per dwelling", "proportion of owner-occupied units built prior to 1940", "weighted mean of distances to five Boston employment centres", "index of accessibility to radial highways", "full-value property-tax rate per \$10,000", "pupil-teacher ratio by town", "1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town", "lower status of the population (percent)" and "median value of owner-occupied homes in \$1000s". This information comes from [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).
  

**The structure and dimensions of the data** are presented here:  

```{r}
data(Boston)
str(Boston)
```


As we can see, the dimensions of Boston data are `r nrow(Boston)` rows and `r ncol(Boston)` columns.
Here is a **summary of the variables**:


```{r}
summary(Boston)
```


Now let's show a graphical overview of the data and show summaries of the variables in the data:

```{r include=FALSE}
library(tidyverse)
library(dplyr)
library(corrplot)
```

```{r}
cor_matrix<-cor(Boston) %>% round(digits = 2)
cor_matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)
```

In the correlation matrix the colors indicate the direction of the correlation (red is negative and blue is positive correlation). Darker colors indicate stronger correlations. Based on this plot we can see that there is one variable pair that has strong positive correlation: "rad" and "tax" ("index of accessibility to radial highways" and "full-value property-tax rate per \$10,000"). Strong negative correlations are found between variables pairs: dis with age, nox and indus; lstat and medv. These variables are:
dis = "weighted mean of distances to five Boston employment centres"
nox = "nitrogen oxides concentration (parts per 10 million)"
indus = "proportion of non-retail business acres per town"
lstat = "lower status of the population (percent)"
medv = "median value of owner-occupied homes in \$1000s"

 
Let's standardize the dataset and print out summaries of the scaled data:

```{r}
boston_scaled <- scale(Boston)
summary(boston_scaled)
```

We can see that after standardizing the mean of each variable is zero. 

```{r include=FALSE}
class(boston_scaled)
boston_scaled <- as.data.frame(boston_scaled)
```

```{r}
# create a quantile vector of crim
bins <- quantile(boston_scaled$crim)

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, label = c("low", "med_low", "med_high", "high"), breaks = bins, include.lowest = TRUE)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]
```

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

```



You can check out my R script for the Data wrangling part [here](https://github.com/MikkoPatronen/IODS-project/blob/master/data/create_human.R).
